Intentional release of native species undermines ecological stability

### 2022-12-23 rmd/manuscript.Rmd
edit: manuscript
anchor: changing to word output

---
title: "Intentional release of native species undermines ecological stability"
output:
  bookdown::pdf_document2:
   latex_engine: xelatex
   toc: false
   number_sections: false
  #bookdown::word_document2:
  # reference_docx: word_format.docx
  # number_sections: false
csl: https://www.zotero.org/styles/pnas
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile,
                          encoding = encoding, 
                          output_dir = "document_output")
      })
bibliography: reference.bib
fontsize: 11pt
mainfont: "Times New Roman"
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

When small carrying capacity was combined with a low to modest intrinsic growth rate ($K=$ `r min(list_param[[2]]$k)`, $r_1 \le 2.5$), our model predicted a destabilizing effect of intentional release on ecological communities, as illustrated by increased CV with increasing numbers of releases (**Figure** \@ref(fig:figure-layout)**A**). This pattern stemmed mainly from the reduced mean of the total community density, and both enhanced and unenhanced species groups were responsible (**Figure** \@ref(fig:figure-layout)**A**). The enhanced species decreased because release induced the negative competitive effect exceeding the reproductive contribution of released individuals, as reported in a previous theoretical study [@satakeStockingCaptivebredFish2012]. Meanwhile, interspecific competition reduced the unenhanced species, resulting in fewer persisting species (**Figure** \@ref(fig:figure-layout)**A**). Combined, the total community density decreased more sharply than individual species groups (**Figure** \@ref(fig:figure-layout)**A**). The SDs showed a similar trend, but the relationship was flatter at the community level (**Figure** \@ref(fig:figure-layout)**A**). Since a CV is a ratio of an SD to a mean, the steeper decline of the mean community density led to the increased CV. These patterns were qualitatively similar across ecological contexts with the low carrying capacity (**Figures S3 -- S6**), but except for the scenario in which the enhanced species exhibits chaos ($r_1=$ `r max(list_param[[2]]$r1)`).

```{r figure-layout, fig.cap= paste0("Theory and empirical observations agree that intentional release destabilizes community dynamics. In A and B, panels and colors distinguish response variables and species groups. (A) Theoretical predictions. Dots represent individual simulation replicates, and lines are the loess curves fitted to simulated data with their 95% confidence (dark shade) and prediction intervals (light shade). Parameters used in this simulation are: intrinsic growth rate of an enhanced species $r_1=$ ", round(na.omit(unique(df_sim$r1)), 2), ", average interspecific competition $\\bar{\\alpha}=$ ", round(na.omit(unique(df_sim$alpha)), 2), ", carrying capacity $K=$ ", round(na.omit(unique(df_sim$k)), 2), ", environmental stochasticity $\\sigma_{\\varepsilon}=$ ", round(na.omit(unique(df_sim$sd_env)), 2), ", relative fitness of captive-bred individuals $f_R=$ ", round(na.omit(unique(df_sim$phi)), 2), ". These parameter values are comparable to those estimated in the empirical Ricker model. (B) Empirical evidence. Dots represent observations at 97 sites within 31 watersheds. Lines and shades are the predicted values and their 95% credible intervals of the regression models, and line types correspond to the coefficient's posterior probabilities. Effective release is a latent explanatory variable estimated in our hierarchical Baysian model. See Methods for details. Full statistics were reported in Tables S6-8. (C) Map of sampling sites (black dots) in Hokkaido, Japan. (D) Example of a protected watershed (Okushibetsu). (E) Masu salmon *Oncorhynchus masou*. Photo credit: Akira Terui")}

knitr::include_graphics(here::here("output/figure_theory_obs.pdf"))

@@

To demonstrate the relevance of our general theory to natural systems, we assessed the potential impacts of the intentional release of masu salmon (**Figure** \@ref(fig:figure-layout)**E**) on the long-term stability of stream fish communities in Hokkaido, Japan. In the protected watersheds (all separated by the ocean; **Figure 1C, D**), a long-term program exists to monitor stream fish communities along with the official release records. The release of masu salmon began in the 1950s, with the annual average ranging from `r round(min(df_stock_mu$mean_stock), 2)` (no release) to `r round(max(df_stock_mu$mean_stock), 2)` million fish across watersheds during 1999 -- 2019. The majority of release occurs in spring near fish survey locations, after which salmon fry stays in freshwater for at least one year with other stream fishes (see **Figure S16** for a co-occurrence matrix in our dataset). Therefore, the study system sets the stage for a "natural experiment" to test our theoretical predictions. We used the data from 1999 to 2019 at `r n_distinct(df_fish$site_id)` sites within `r n_distinct(df_fish$river)` independent watersheds (see **Methods** for selection criteria). Using hierarchical Bayesian models, we quantified the effect of intentional release on community dynamics while accounting for the potential effects of climates and local abiotic factors.

As predicted, stream fish communities showed greater temporal fluctuations (higher CV) at sites with the intensive release (**Figure** \@ref(fig:figure-layout)**B**). The effect was striking in its magnitude, almost doubling the CV at the highest release level. Our analysis strongly supported the positive relationship between the CV and the number of "effective" releases (**Figure** \@ref(fig:figure-coef); see **Methods** for the Bayesian estimation of effective release), in which the probability of the regression coefficient being positive was `r op(df_beta %>% filter(response == "cv") %>% pull(prob) %>% round(2), 2)` (**Table S6**). This pattern was associated with the reduced long-term average of the total community density and fewer taxonomic richness (**Figures** \@ref(fig:figure-layout)**B and** \@ref(fig:figure-coef)), and both enhanced (masu salmon) and unenhanced fish groups contributed to this trend (**Figures** \@ref(fig:figure-layout)**B and** \@ref(fig:figure-coef)**;** see **Tables S6 -- S8** for full statistics). In the meantime, the SDs had vague relationships with the intentional release (**Figures** \@ref(fig:figure-layout)**B and** \@ref(fig:figure-coef)).

```{r figure-coef, fig.cap="Posterior distributions for the standardized regression coefficients of intentional release. Y-axis represents different response variables grouped by colors distinguishing species groups (whole community, enhanced species [masu salmon], and unenhanced species). Triangles indicate median estimates."}

knitr::include_graphics(here::here("output/figure_coef.pdf"))

To gain better insights into the driving mechanisms, we fitted a multi-species Ricker model to a subset of our time-series data (nine sites with few missing observations and no intentional release; **Methods**). The parameter estimates of this model are directly comparable to those used in the theoretical Ricker model. The estimates of interspecific competition coefficients $\alpha_{ij}$ ranged `r op(min(df_pd$alpha_prime), 2)` -- `r op(max(df_pd$alpha_prime), 2)` with the median of `r op(median(df_pd$alpha_prime), 2)` (**Figure** \@ref(fig:figure-alpha)), suggesting that competition is a likely mechanism as our theory assumes (see **Figure S17** for all pairwise estimates). Our statistical inference is supported by previous studies. For example, competition between masu salmon and other salmonid species is well documented, and its form can be interference or exploitative [@miyasakaFoodHabitDivergence2003; @hasegawaComparisonCompetitiveAbility2004; @moritaImpactExoticTrout2004]. Diets of salmonid species are innately overlapped, and they interfere with each other to claim profitable foraging spots for drift insects [@hasegawaComparisonCompetitiveAbility2004] or cruise over a large area in search of prey [@hasegawaEffectFlowRegime2010]. Exploitation competition between masu salmon and benthic fish (e.g., sculpins and stone loaches in our study system) is also likely because subordinate masu salmon in the dominance hierarchy consumes benthic macroinvertebrates [@satoStagespecificFunctionalResponses2014] that many benthic fish species prey on [@natsumedaEvaluationEcologicalFeatures2010; @natsumedaDietFoodSelectivity2012]. Importantly, hatchery masu salmon may intensify intra- and interspecific competition in the wild [@hasegawaWildMasuSalmon2018; @sahashiEvaluationContributionHatchery2015; @hasegawaPulsedSuppliesSmall2022]. Hatchery salmons are larger and more aggressive than wild individuals of a comparable life stage, increasing the likelihood of intense intra- and interspecific competition [@weberInteractionsHatcheryWild2003]. Indeed, field and experimental studies confirmed that hatchery masu salmon competed with wild masu salmon and other stream fishes during their freshwater life stage [@hasegawaWildMasuSalmon2018; @sahashiEvaluationContributionHatchery2015; @hasegawaPulsedSuppliesSmall2022].

@@

```{r figure-alpha, fig.cap="Estimated competition coefficients of the multi-species Ricker model. Dots represent a given pair of taxa, and box plots show median (vertical center line) and quartiles (box limits) with whiskers extending up to the most extreme data points that are within $\\pm 1.5$ interquartile range. The density plot on the top shows the overall distribution of the competition coefficient. Colors distinguish taxa. The vertical broken line denotes the overall median value."}

source(here::here("code/si_figure_species_int.R"))

knitr::include_graphics(here::here("output/figure_alpha_dist.pdf"))
```

@@

>>>>

---
title: "Intentional release of native species undermines ecological stability"
output:
  # bookdown::pdf_document2:
  #  latex_engine: xelatex
  #  toc: false
  #  number_sections: false
  bookdown::word_document2:
    reference_docx: word_format.docx
    number_sections: false
    pandoc_args: ["-Fpandoc-crossref"]
csl: https://www.zotero.org/styles/pnas
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile,
                          encoding = encoding, 
                          output_dir = "document_output")
      })
bibliography: reference.bib
fontsize: 11pt
mainfont: "Times New Roman"
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

When small carrying capacity was combined with a low to modest intrinsic growth rate ($K=$ `r min(list_param[[2]]$k)`, $r_1 \le 2.5$), our model predicted a destabilizing effect of intentional release on ecological communities, as illustrated by increased CV with increasing numbers of releases (**Figure** 1)**A**). This pattern stemmed mainly from the reduced mean of the total community density, and both enhanced and unenhanced species groups were responsible (**Figure** 1)**A**). The enhanced species decreased because release induced the negative competitive effect exceeding the reproductive contribution of released individuals, as reported in a previous theoretical study [@satakeStockingCaptivebredFish2012]. Meanwhile, interspecific competition reduced the unenhanced species, resulting in fewer persisting species (**Figure** 1)**A**). Combined, the total community density decreased more sharply than individual species groups (**Figure** 1)**A**). The SDs showed a similar trend, but the relationship was flatter at the community level (**Figure** 1)**A**). Since a CV is a ratio of an SD to a mean, the steeper decline of the mean community density led to the increased CV. These patterns were qualitatively similar across ecological contexts with the low carrying capacity (**Figures S3 -- S6**), but except for the scenario in which the enhanced species exhibits chaos ($r_1=$ `r max(list_param[[2]]$r1)`).

@@

To demonstrate the relevance of our general theory to natural systems, we assessed the potential impacts of the intentional release of masu salmon (**Figure** 1)**E**) on the long-term stability of stream fish communities in Hokkaido, Japan. In the protected watersheds (all separated by the ocean; **Figure 1C, D**), a long-term program exists to monitor stream fish communities along with the official release records. The release of masu salmon began in the 1950s, with the annual average ranging from `r round(min(df_stock_mu$mean_stock), 2)` (no release) to `r round(max(df_stock_mu$mean_stock), 2)` million fish across watersheds during 1999 -- 2019. The majority of release occurs in spring near fish survey locations, after which salmon fry stays in freshwater for at least one year with other stream fishes (see **Figure S16** for a co-occurrence matrix in our dataset). Therefore, the study system sets the stage for a "natural experiment" to test our theoretical predictions. We used the data from 1999 to 2019 at `r n_distinct(df_fish$site_id)` sites within `r n_distinct(df_fish$river)` independent watersheds (see **Methods** for selection criteria). Using hierarchical Bayesian models, we quantified the effect of intentional release on community dynamics while accounting for the potential effects of climates and local abiotic factors.

As predicted, stream fish communities showed greater temporal fluctuations (higher CV) at sites with the intensive release (**Figure** 1)**B**). The effect was striking in its magnitude, almost doubling the CV at the highest release level. Our analysis strongly supported the positive relationship between the CV and the number of "effective" releases (**Figure** 2; see **Methods** for the Bayesian estimation of effective release), in which the probability of the regression coefficient being positive was `r op(df_beta %>% filter(response == "cv") %>% pull(prob) %>% round(2), 2)` (**Table S6**). This pattern was associated with the reduced long-term average of the total community density and fewer taxonomic richness (**Figures** 1)**B and** 2), and both enhanced (masu salmon) and unenhanced fish groups contributed to this trend (**Figures** 1)**B and** 2**;** see **Tables S6 -- S8** for full statistics). In the meantime, the SDs had vague relationships with the intentional release (**Figures** 1)**B and** 2).

To gain better insights into the driving mechanisms, we fitted a multi-species Ricker model to a subset of our time-series data (nine sites with few missing observations and no intentional release; **Methods**). The parameter estimates of this model are directly comparable to those used in the theoretical Ricker model. The estimates of interspecific competition coefficients $\alpha_{ij}$ ranged `r op(min(df_pd$alpha_prime), 2)` -- `r op(max(df_pd$alpha_prime), 2)` with the median of `r op(median(df_pd$alpha_prime), 2)` (**Figure** 3)), suggesting that competition is a likely mechanism as our theory assumes (see **Figure S17** for all pairwise estimates). Our statistical inference is supported by previous studies. For example, competition between masu salmon and other salmonid species is well documented, and its form can be interference or exploitative [@miyasakaFoodHabitDivergence2003; @hasegawaComparisonCompetitiveAbility2004; @moritaImpactExoticTrout2004]. Diets of salmonid species are innately overlapped, and they interfere with each other to claim profitable foraging spots for drift insects [@hasegawaComparisonCompetitiveAbility2004] or cruise over a large area in search of prey [@hasegawaEffectFlowRegime2010]. Exploitation competition between masu salmon and benthic fish (e.g., sculpins and stone loaches in our study system) is also likely because subordinate masu salmon in the dominance hierarchy consumes benthic macroinvertebrates [@satoStagespecificFunctionalResponses2014] that many benthic fish species prey on [@natsumedaEvaluationEcologicalFeatures2010; @natsumedaDietFoodSelectivity2012]. Importantly, hatchery masu salmon may intensify intra- and interspecific competition in the wild [@hasegawaWildMasuSalmon2018; @sahashiEvaluationContributionHatchery2015; @hasegawaPulsedSuppliesSmall2022]. Hatchery salmon are larger and more aggressive than wild individuals of a comparable life stage, increasing the likelihood of intense intra- and interspecific competition [@weberInteractionsHatcheryWild2003]. Indeed, field and experimental studies confirmed that hatchery masu salmon competed with wild masu salmon and other stream fishes during their freshwater life stage [@hasegawaWildMasuSalmon2018; @sahashiEvaluationContributionHatchery2015; @hasegawaPulsedSuppliesSmall2022].

@@

@@

$$N_{i,t+1} = N_{i,t}~\exp\left[r_i \left(1-\frac{\sum_{j=1}^S \alpha_{ij}N_{j,t}}{K_i} \right) \right]\exp(\varepsilon_{i,t})$${#eq:ricker0}
